---
title: "Foreasting Time Series Data in R with tidy"
author: "Badawi Amin Muhammed"
date: "`r Sys.Date()`"
output: html_document
theme: flatly #CLEAN VISUAL STYLE
code: folding 
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  echo=TRUE, #shows code
  warning=FALSE, # hide warnings
  message=FALSE,# hide package message
  fig.align = "center", # center all plots
  fig.width = 7 ,#default width
  fig.height = 5, #default height
  out.extra = "style='border:2px solid #444; 
                padding:6px;
                border-radius:8px;
                background:white'"
)
```


# Introduction

For almost a decade, the `forecast` package has been a rock-solid framework for time series forecasting. However, within the last year or so an official updated version has been released named `fable` which now follows `tidy` methods as opposed to base R.

More recently, `modeltime` has been released and this also follows tidy methods. However, it is strictly used for modeling. For data manipulation and visualization, the `timetk` package will be used which is written by the same author as `modeltime`.

The following is a code comparison of various time series visualizations between these frameworks: `fpp2`, `fpp3` and `timetk`.

#Load Libraries
```{r load libraries, message = FALSE, warning = FALSE}
# Load libraries
library(fpp2)         # An older forecasting framework
library(fpp3)         # A newer tidy forecasting framework
library(timetk)       # An even newer tidy forecasting framework
library(tidyverse)    # Collection of data manipulation tools
library(tidyquant)    # Business Science ggplot theme
library(cowplot)      # A ggplot add-on for arranging plots
```

# TS vs tsibble
- the base *ts* object is used by `forecast` & `fpp2`
- The special *tsibble*  object is used by `fable` & `fpp3`
- The standard tibble object is used by `timetk` & `modeltime`

# 1. Load Time Series Data
we will utilize a dataset containing quarterly production values of certain commodities in Australia.
```{r loading data, message = FALSE, warning = FALSE}
# Quarterly Australian production data as tibble
aus <- tsibbledata::aus_production %>% as_tibble()
# Check structure
aus %>% str()
```

# 2 fpp2 Method: From tibble to ts
```{r converting to ts, message = FALSE, warning = FALSE}
# Convert tibble to time series object
aus_prod_ts <- ts(aus[, 2:7],  # Choose columns
                    start = c(1956, 1),  # Choose start date
                    end = c(2010, 2),    # Choose end date
                    frequency = 4)       # Choose frequency per yr
# Check it out
aus_prod_ts %>% tail()
```

#3 fpp3 Method: From ts to tsibble

## 3.1 Pivot Wide

```{r keeping wide data format, message = FALSE, warning = FALSE}
# Convert ts to tsibble and keep wide format
aus_prod_tbl_wide <- aus_prod_ts %>%    # TS object
  as_tsibble(index = "index",           # Set index column
             pivot_longer = FALSE)      # Wide format
```

 ## 3.2 Pivot Long
      Convert ts to tsibble and pivot to long format
 
```{r Convert tS and pivot to long format,  message = FALSE, warning = FALSE}
aus_prod_tbl_long <- aus_prod_ts %>%    # TS object
  as_tsibble(index = "index", pivot_longer = TRUE) # Long format
```

#4 timetk Method: From tsibble/ts to tibble
## 4.1 Pivot Wide

```{r keeping wide format, message = FALSE, warning = FALSE}
# Convert tsibble to tibble, keep wide format
aus <- tsibbledata::aus_production %>% 
    tk_tbl() %>%
    mutate(Quarter = as_date(as.POSIXct.Date(Quarter)))
```

## 4.2 Pivot Long
```{r quarterly production data, message = FALSE, warning = FALSE}
# Quarterly Australian production data to long format
aus_long <- aus %>% 
    rename(date = Quarter) %>%
    pivot_longer(
        cols = c("Beer","Tobacco","Bricks",
                 "Cement","Electricity","Gas"))
```


# 5. Time Series Plots
When analyzing time series plots, look for the following patterns:

-**Trend**: A long-term increase or decrease in the data; a “changing direction”.

_**Seasonality**: A seasonal pattern of a fixed and known period. If the frequency is unchanging and associated with some aspect of the calendar, then the pattern is seasonal.

-**Cycle**: A rise and fall pattern not of a fixed frequency. If the fluctuations are not of a fixed frequency then they are cyclic.

-**Seasonal vs Cyclic**: Cyclic patterns are longer and more variable than seasonal patterns in general.

# 5.1 fpp2 Method: Plot Multiple Series On Same Axes
```{r Multiple series on thesame axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using fpp2"}
# Using fpp2
aus_prod_ts %>%               # TS object
  autoplot(facets=FALSE)      # No facetting
```

## 5.2 fpp3 Method: Plot Multiple Series On Same Axes

```{r Multiple series on x axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using fpp3" }
# Using fpp3
aus_prod_tbl_long %>%    # Data in long format
  autoplot(value) 
```
## 5.3 ggplot Method: Plot Multiple Series On Same Axes
plotting multiple plots on the same axes has not been implemented into `timetk`. hence, we Use `ggplot`.

```{r  Multiple series on same axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using ggplot" }
# Using ggplot
aus_long %>%
    ggplot(aes(date, value, group = name, color = name)) +
    geom_line()
```

## 5.4 fpp2 Method: Plot Multiple Series On Separate Axes
```{r Multiple series on fpp2 axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using fpp2"}
# Using fpp2
aus_prod_ts %>%  
  autoplot(facets=TRUE)   # With facetting
```

## 5.5 fpp3 Method: Plot Multiple Series On Separate Axes
```{r  Multiple series on fpp3 axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using fpp3"}
# Using fpp3
aus_prod_tbl_long %>%
  ggplot(aes(x = index, y = value, group = key)) + 
  geom_line() + 
  facet_grid(vars(key), scales = "free_y")   # With facetting
```

## 5.6 timetk Method: Plot Multiple Series On Separate Axes
```{r  Multiple series on timetk axis, warning=FALSE, message=FALSE, fig.cap="quaterly production of selected quantities using timetk"}
# Using timetk
aus_long %>% 
    plot_time_series(
        .date_var = date,
        .value = value,
        .facet_vars = c(name), # Group by these columns
        .color_var = name, 
        .interactive = FALSE,
        .legend_show = FALSE
    )
```

# Seasonal Plots
Use seasonal plots for identifying time periods in which the patterns change.

 ## 6.1 fpp2 Method: Plot Individual Seasons
```{r fpp2 method, warning=FALSE, message=FALSE, fig.cap="Monthly plot of anti-diabetic scripts"}
# Monthly plot of anti-diabetic scripts in Australia 
a1 <- a10 %>%
  autoplot() 
# Seasonal plot
a2 <- a10 %>% 
  ggseasonplot(year.labels.left = TRUE,   # Add labels
               year.labels = TRUE) 
# Arrangement of plots
plot_grid(a1, a2, ncol=1, rel_heights = c(1, 1.5))
```

# 6.2 fpp3 Method: Plot Individual Seasons
```{r fpp3 method, warning=FALSE, message=FALSE, fig.cap="Monthly plot of anti-diabetic scripts"}
# Monthly plot of anti-diabetic scripts in Australia
a1 <- a10 %>%
  as_tsibble() %>%
  autoplot(value)
# Seasonal plot
a2 <- a10 %>%
  as_tsibble() %>%
  gg_season(value, labels="both")   # Add labels
# Arrangement of plots
plot_grid(a1, a2, ncol=1, rel_heights = c(1, 1.5))
```

## 6.3 ggplot Method: Plot Individual Seasons

Note that seasonal plots have not been implemented into `timetk` .hence, Use `ggplot` to write:
```{r ggplot method, warning=FALSE, message=FALSE, fig.cap="Monthly plot of anti-diabetic scripts"}
# Convert ts to tibble
a10_tbl <- fpp2::a10 %>%
    tk_tbl()
# Monthly plot of anti-diabetic scripts in Australia 
a1 <- a10_tbl %>% 
    plot_time_series(
        .date_var = index,
        .value    = value,
        .smooth   = TRUE,
        .interactive = FALSE,
        .title = "Monthly anti-diabetic scripts in Australia"
    )
# New time-based features to group by
a10_tbl_add <- a10_tbl %>% 
    mutate( 
        month = factor(month(index, label = TRUE)),  # Plot this
        year = factor(year(index))  # Grouped on y-axis
    )
# Seasonal plot
a2 <- a10_tbl_add %>%
    ggplot(aes(x = month, y = value, 
               group = year, color = year)) + 
    geom_line() + 
    geom_text(
        data = a10_tbl_add %>% filter(month == min(month)),
        aes(label = year, x = month, y = value),
        nudge_x = -0.3) + 
    geom_text(
        data = a10_tbl_add %>% filter(month == max(month)),
        aes(label = year, x = month, y = value),
        nudge_x = 0.3) + 
    guides(color = FALSE)
# Arrangement of plots
plot_grid(a1, a2, ncol=1, rel_heights = c(1, 1.5))
```
# 7. Lag Plots

Use lag plots to check for randomness.

## 7.1 fpp2 Method: Plot Multiple Lags
```{r fpp2 multiple lags, warning=FALSE, message=FALSE, fig.cap="Annual oil production in Saudi Arabia"}
# Plot of non-seasonal oil production in Saudi Arabia
o1 <- fpp2::oil %>%
  autoplot()
# Lag plot of non-seasonal oil production
o2 <- gglagplot(oil, do.lines = FALSE)
# Plot both
plot_grid(o1, o2, ncol=1, rel_heights = c(1,2))
```
## 7.2 fpp3 Method: Plot Multiple Lags

```{r fpp3 multiple lags, warning=FALSE, message=FALSE, fig.cap="Annual oil production in Saudi Arabia"}
# Plot of non-seasonal oil production
o1 <- oil %>%
  as_tsibble() %>%
  autoplot(value)
# Lag plot of non-seasonal oil production
o2 <- oil %>%
  as_tsibble() %>%
  gg_lag(y=value, geom = "point") 
# Plot it
plot_grid(o1, o2, ncol=1, rel_heights = c(1,2))
```


## 7.3 timetk Method: Plot Multiple Lags

```{r convert to tibble first, warning=FALSE, message=FALSE}
# Convert to tibble and create lag columns
oil_lag_long <- oil %>%
    tk_tbl(rename_index = "year") %>%
    tk_augment_lags(      # Add 9 lag columns of data
        .value = value, 
        .names = "auto", 
        .lags = 1:9) %>%
    pivot_longer(         # Pivot from wide to long
        names_to = "lag_id", 
        values_to = "lag_value", 
        cols = value_lag1:value_lag9) # Exclude year & value
```

then the plot:
```{r warning=FALSE, message=FALSE, fig.cap="Annual oil production in Saudi Arabia"}
# Time series plot
o1 <- oil %>%
    tk_tbl(rename_index = "year") %>%  
    mutate(year = ymd(year, truncated = 2L)) %>%  
    plot_time_series(
        .date_var = year, 
        .value = value,
        .interactive = FALSE)
# timetk Method: Plot Multiple Lags
o2 <- oil_lag_long %>%
    plot_time_series(
        .date_var = value,     # Use value instead of date
        .value = lag_value,    # Use lag value to plot against
        .facet_vars = lag_id,  # Facet by lag number
        .facet_ncol = 3,
        .interactive = FALSE, 
        .smooth = FALSE,
        .line_alpha = 0,      
        .legend_show = FALSE,
        .facet_scales = "fixed"
        ) + 
    geom_point(aes(colour = lag_id)) + 
    geom_abline(colour = "gray", linetype = "dashed") 
# Plot it
plot_grid(o1, o2, ncol=1, rel_heights = c(1,2))
```
# 8. Autocorrelation Function Plots

The autocorrelation function measures the linear relationship between lagged values of a time series. The partial autocorrelation function measures the linear relationship between the correlations of the residuals.

**ACF**

-Visualizes how much the most recent value of the series is correlated with past values of the series (lags)
-If the data has a trend, then the autocorrelations for small lags tend to be positive and large because observations nearby in time are also nearby in size
-If the data are seasonal, then the autocorrelations will be larger for seasonal lags at multiples of seasonal frequency than other lags

**PACF**

-Visualizes whether certain lags are good for modeling or not; useful for data with a seasonal pattern
Removes dependence of lags on other lags by using the correlations of the residuals

## 8.1 fpp2 Method: Plot ACF + PACF
```{r ACF PLOT using fpp2, warning=FALSE, message=FALSE, fig.cap="Auto Correlation Function" }
# ACF plot 
o1 <- ggAcf(oil, lag.max = 20)
# PACF plot
o2 <- ggPacf(oil, lag.max = 20)
# Plot both
plot_grid(o1, o2, ncol = 1)
```


## 8.2 fpp3 Method: Plot ACF + PACF

```{r ACF PLOT using fpp3, warning=FALSE, message=FALSE, fig.cap="Auto Correlation Function"}
# Convert to tsibble
oil_tsbl <- oil %>% as_tsibble()
# ACF Plot
o1 <- oil_tsbl %>%
  ACF(lag_max = 20) %>%
  autoplot()
# PACF Plot
o2 <- oil_tsbl %>%
  PACF(lag_max = 20) %>%
  autoplot() 
# Plot both
plot_grid(o1, o2, ncol = 1)
```

## 7.3 timetk Method: Plot ACF & PACF

```{r ACF PLOT using timetk, warning=FALSE, message=FALSE, fig.cap="Auto Correlation Function"}
# Using timetk
oil %>%
    tk_tbl(rename_index = "year") %>%
    plot_acf_diagnostics(
        .date_var = year,
        .value    = value,
        .lags     = 20,
        .show_white_noise_bars = TRUE, 
        .interactive = FALSE
    )
```

# Summary

  As with all things in life, there are good and bad sides to using any of these three forecasting frameworks for visualizing time series. All three have similar functionality as it relates to visualizations.

 **fpp2**
 
-Code requires minimal parameters
-Uses basets format
-Uses `ggplot` for visualizations
-Mostly incompatible with tidyverse for data manipulation
-No longer maintained except for bug fixes

 **fpp3**
 
-Code requires minimal parameters
-Uses proprietary tsibble format with special indexing tools
-Uses `ggplot` for visualizations
-Mostly compatible with tidyverse for data manipulation; tsibble may cause issues
-Currently maintained

  **timetk**
  
-Code requires multiple parameters but provides more granularity
-Uses standard tibble format
-Uses `ggplot` and `plotly` for visualizations
-Fully compatible with `tidyverse` for data manipulation
-Currently maintained


